---
title: "The `ggtufte` Package"
author: "Jeffrey Arnold"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(tufte)
# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, cache.extra = packageVersion('tufte'))
options(htmltools.dir.version = FALSE)
```

# Introduction

```{r}
library("ggplot2")
library("ggtufte")
library("scales")
extrafont::loadfonts(quiet = TRUE)

```

## Scatter plots

```{r}
p <- ggplot(mtcars, aes(wt, mpg)) +
  geom_point() 
```

```{r}
p + theme_bw()
```

The **ggplot2** theme provided with this package, `theme_tufte`, removes most of the
ink from the plot. There are no grid-lines and no borders or axis.
```{r}
p + theme_tufte() 
```

Instead of using axis lines the ink-to-data ratio can be improved by using the axis lines
to encode the range of the data.
The `geom_rangeframe()` will add lines representing the range of the data along the axes.
```{r}
p + theme_tufte() +
  geom_rangeframe()
```

However, the axis breaks do not match up with the range. 
It would be more informative to label the minimum and maximum values of breaks.
```{r}
p + theme_tufte() +
  geom_rangeframe() +
  scale_y_continuous(breaks = range_breaks(),
                     labels = format_format(digits = 1)) +
  scale_x_continuous(breaks = range_breaks(),
                     labels = format_format(digits = 1))
```

```{r}
p + theme_tufte() +
  geom_rangeframe() +
  scale_y_continuous(breaks = range_breaks()(mtcars$mpg),
                     labels = format_format(digits = 1)) +
  scale_x_continuous(breaks = range_breaks()(mtcars$wt),
                     labels = format_format(digits = 1))
```

```{r}
p + theme_tufte(ticks = FALSE) +
  geom_rug() +
  scale_y_continuous(breaks = range_breaks()(mtcars$mpg),
                     labels = format_format(digits = 1)) +
  scale_x_continuous(breaks = range_breaks()(mtcars$wt),
                     labels = format_format(digits = 1))
```

## Quartile Plots

```{r}
p <- ggplot(mpg, aes(class, hwy)) +
  theme(axis.ticks.x = element_blank())
```

Tufte Ch. 6 proposes several redesigns of the venerable box plot.
Unlike the Tukey's boxplot, these box plots, display the five-number summary
(median, 25% quartile, 75% quartile, maximum, minimum), instead of the 
stats and outliers calculated in Tukey's boxplot.

This displays the standard box plot in **ggplot2**.
```{r}
p + geom_boxplot(stat = StatFivenumber)
```
I use the `StatFivenumber` stat to calculate the fiver-number summary rather
than the summary used in the boxplot.

@Tufte calls his redesign a *quartile-plot*. 
The function `geom_quartileplot()` implements these.

The first redesign erases the box from the boxplot.
The whiskers remain, negative space represents the interquartile and a point represents the median.
```{r}
p + geom_quartileplot()
```

A second alternative is to invert the first.
Points are used to represent the minimum and maximum, a line for the interquartile
range, and a gap in the line to represent the median.
```{r}
p + geom_quartileplot(whisker.line = FALSE, middle.line = TRUE)
```

A third alternative is to use a line to represent the range, and a gap to 
represent the median.
The interquartile range is represented by a thicker section of line.
```{r}
p + geom_quartileplot(whisker.line = TRUE, middle.line = TRUE,
                      hoffset = 0, fatten = 2)
```

A fourth alternative uses lines for the upper and lower whiskers.
The interquartile range is represented by a line offset from the first.
The median is still represented by a gap.
```{r}
p + geom_quartileplot(whisker.line = TRUE, middle.line = TRUE)
```

## Theme

```{r}
p <- ggplot(mtcars, aes(x = wt, y = mpg, colour = factor(cyl))) +
  geom_point() + 
  facet_wrap(~ am) +
  labs(title = "Title", subtitle = "Subtitle", caption = "Caption")
p + theme_tufte()
```

```{r}
p + theme_tufte(sans_title = TRUE)
```


## Typography



## References

- [Tufte Handout](https://rstudio.github.io/tufte/) R markdown format
- [tufte-css](https://github.com/edwardtufte/tufte-css)
- [tufte-latex](https://github.com/tufte-latex/tufte-latex)
